\textbf{Real-time guitar transcription using\\Fourier transform based methods}\\A pitch estimation framework and overtone sieve algorithm

### 2021-04-15 paper.tex
foutjes

The resolution of the Fourier transform is dependent on the number of samples in a frame. We sample the guitar signal at 192 kHz, which is the maximum sample rate for most audio interfaces. To allow guitarists to play mildly fast, we have to limit our frame length. If the frame size if too large, multiple separate notes may be included in one transform, which would make it seem like the notes were played at the same time. Moreover, frame sizes which are a power of two can be transformed more efficiently, which further restricts are frame size choice. We chose a frame size of $2^{14} = 16384$ samples, which at 192 kHz is 0.085 seconds (11.72 frames per second). This gives us a Fourier bin size of 11.72 Hz. At first glance, such a large bin size seems like a big problem, as the smallest interval that can be played on a guitar (G\#\textsubscript{2}-A\textsubscript{2}) is about 6 Hz. However, the overtones of these notes will have more Hz between them, which makes to possible to differentiate the two notes.% Also, as discussed in Section~\ref{sec:future}, we might be able to increase the frame size without reducing our frame rate too much.

@@

We constructed our own dataset to evaluate the performance. This dataset is a technical dataset consisting of scales and intervals played on a guitar. Scales allow us to measure the monophonic performance. As scales are easy to play, multiple recordings should have the same results. This allows us to isolate and measure other variables, such as the effect of tempo or playing the scale higher/lower. Recordings of different intervals allow us to measure the polyphonic performance. By measuring all two note intervals, we can identify any problems with specific intervals. As the distance in frequency in an interval increases when the same interval is played at higher notes, we record the intervals at multiple root notes. We will also measure the performance on the four basic triads: the major triad, minor triad, augmented triad and diminished triad.

@@

\subsection{Monophonic performance: scales}
To test the monophonic performance of our system, we first measure the performance of scales at different pitch ranges. The scales are played at a speed of two notes per second. This equates to playing fourth notes on 120 BPM. We play the scales at a relatively slow speed, such that this cannot become a limiting factor. The slower speed also helps to see if mistakes are the result of transients, as these mistakes should vanish as the note persists. We only play the C major scale, but change the starting note (mode). We play the full vertical scale from this starting note~\cite{scale}. Each position has 17 notes, except for $\text{F}_2$, which has 18 notes. In Table~\ref{tab:monperf}, we summarize the results of the experiment. We denote the different starting positions as $n_\text{start}$. For every starting position, we denote the accuracy ($a$), the accuracy without transient errors ($a_\text{trans}$), the accuracy without octave errors $a_\text{oct}$ and the adjusted accuracy without both of these errors ($a_\text{adj}$). The accuracy is calculated by dividing the number of correctly determined frames by the total number of frames with notes. The only frames without notes are positioned at the start and end of a file. We omitted these frames from the results, as they are always correctly determined to be empty frames. Erroneous notes due to transients usually have a lower amplitude where octave errors usually have a higher amplitude compared the amplitude of the base note. This allows is to distinguish between them. This information cannot be used to detect transients in real-time, as we compare the amplitude of the erroneous note to amplitude of the correct note in the following frames.
\begin{table}[H]\vspace{-3mm}
    \centering
    \begin{tabular}{r|rrrr}
        $n_\text{start}$ & $a$ & $a_\text{trans}$ & $a_\text{oct}$ & $a_\text{adj}$ \\
        \hline
        $\text{E}_2$ & 91.4\% & 95.2\% & 96.2\% & 100\% \\
        $\text{F}_2$ & 92.3\% & 96.2\% & 96.2\% & 100\% \\
        $\text{G}_2$ & 92.0\% & 95.0\% & 97.0\% & 100\% \\
        $\text{A}_2$ & 93.0\% & 94.0\% & 99.0\% & 100\% \\
        $\text{B}_2$ & 92.8\% & 96.9\% & 95.9\% & 100\% \\
        $\text{C}_3$ & 92.8\% & 92.8\% & 100\%  & 100\% \\
        $\text{D}_3$ & 96.0\% & 98.0\% & 98.0\% & 100\% \\
        $\text{E}_3$ & 96.7\% & 96.7\% & 100\%  & 100\%
    \end{tabular}
    \caption{Accuracy of monophonic $f_0$ estimation on C major scale at two notes per second for different starting positions. $a_\text{trans}$ is the accuracy without transient errors and $a_\text{oct}$ is without octave errors. $a_\text{adj}$ is the accuracy without both of these errors.}
    \label{tab:monperf}
\end{table}\vspace{-3mm}
We can see that the transient errors occur about 96\% of the time. There are two outliers at $\text{C}_3$ and $\text{D}_3$, which are the cause of inconsistent quality the played scales. This shows us that the guitarist has to play carefully for the best results. Furthermore, we noticed that the first played note almost always had a transient error. This could be because we start playing from a silent signal instead of changing from a previous note. Also, the transient errors show up more often between lower frequency notes. Almost all octave errors are at note changes, so better transient filtering will also solve these cases. Other than these errors, the system seems to work perfectly.

@@

To test the polyphonic performance of our system, we let it determine the played notes in an interval. We play every interval within an octave to test if there are any problems with separating the notes in specific intervals. For instance, some interference may cause oscillating peaks in the frequency domain. Again, we try all the intervals at different root notes, as the pitch of the root note may make a big difference in performance. As we mentioned before, when two peaks are too close together, we cannot distinguish them. Often, the second note is distinguished an octave higher. We set the overtone error threshold to 10 cents. This is a relatively small value, but this allows us to see small problems in our algorithms better. Furthermore, as described in~\ref{sec:future}, we should not use a constant value, but use a different value for every $n$-th overtone. In Tables~\ref{tab:polperf1}--\ref{tab:polperf3} is a summary of the results. Here, we list the number of semitones $n$ between the two notes in an interval, if both notes were correctly found. If not, we denote if the second note was detected an octave higher. This often happens, because we cannot distinguish two notes in small intervals at low pitches. The last column shows if there were other notes reported which were not transients.

@@

\section{Conclusions}
In this paper, we presented and implemented a Fourier transform based AMT system and tested if these methods are feasible for real-time transcription. We divided the task in three components which can all be separately interchanged to facility specialized research in one of the components. The components are: obtaining the frequency domain representation, spectral peak picking and note selection. As shown in Section~\ref{sub:algspeed}, these tasks can be performed well within our real-time constraint of 6 ms. However, when using Fourier transform based methods, we need large frame sizes to get the required accuracy to discern lower pitched notes. Given the maximum sampling rate of widespread audio interfaces, it takes too long to fill a frame. The current implementation requires frame sizes of 16384 samples, which equates to frame lengths of 85 ms. As described in~\ref{sec:future}, we might still be able to use Fourier transform based methods for real-time transcription. However, other transforms such as CQT or wavelet transform might be better suited for real-time transcription, as we can still spare much compute time to lower the frame-wait time.

@@

The current implementation has no onset detection. We stated before that onset detection may help to filter out transients. If we can detect the transient in the time domain, we can discard all samples containing the transient. This may however be very difficult to do in real-time. If we can detect that there was a transient in a transformed frame, we are too late to act on it due to the large frame times. We can circumvent this by transforming partially overlapping frames. This could be done in different threads to limit the extra computational pressure which could impact our latency. However, as we are about 240 times real-time, we could likely perform it single threaded. Now, if we have a bad frame, we have to wait a full frame (83 ms) for the next estimation. However, if we use $n$ partially overlapping frames, we can try another estimation in $83/n$ ms. This will increase the responsiveness of our system greatly.

>>>>

The resolution of the Fourier transform is dependent on the number of samples in a frame. We sample the guitar signal at 192 kHz, which is the maximum sample rate for most audio interfaces. To allow guitarists to play mildly fast, we have to limit our frame length. If the frame size if too large, multiple separate notes may be included in one transform, which would make it seem like the notes were played at the same time. Moreover, frame sizes which are a power of two can be transformed more efficiently, which further restricts are frame size choice. We chose a frame size of $2^{14} = 16384$ samples, which at 192 kHz is 85.33 ms (11.72 frames per second). This gives us a Fourier bin size of 11.72 Hz. At first glance, such a large bin size seems like a big problem, as the smallest interval that can be played on a guitar (G\#\textsubscript{2}-A\textsubscript{2}) is about 6 Hz. However, the overtones of these notes will have more Hz between them, which makes to possible to differentiate the two notes.% Also, as discussed in Section~\ref{sec:future}, we might be able to increase the frame size without reducing our frame rate too much.

@@

We constructed our own dataset to evaluate the performance. This dataset is a technical dataset consisting of scales and intervals played on a guitar. Scales allow us to measure the monophonic performance. As scales are easy to play, multiple recordings should have the same results. This allows us to isolate and measure other variables, such as the effect of tempo or playing the scale higher/lower. Recordings of different intervals allow us to measure the polyphonic performance. By measuring all two note intervals, we can identify any problems with specific intervals. As the distance in frequency in an interval increases when the same interval is played at higher notes, we record the intervals at multiple root notes.% We will also measure the performance on the four basic triads: the major triad, minor triad, augmented triad and diminished triad.

@@

\subsection{Monophonic performance: scales}
To test the monophonic performance of our system, we first measure the performance of scales at different pitch ranges. The scales are played at a speed of two notes per second. This equates to playing fourth notes on 120 BPM. We play the scales at a relatively slow speed, such that this cannot become a limiting factor. The slower speed also helps to see if mistakes are the result of transients, as these mistakes should vanish as the note persists. We only play the C major scale, but change the starting note (mode). We play the full vertical scale from this starting note~\cite{scale}. Each position has 17 notes, except for $\text{F}_2$, which has 18 notes. In Table~\ref{tab:monperf}, we summarize the results of the experiment. We denote the different starting positions as $n_\text{start}$. For every starting position, we denote the accuracy ($a$), the accuracy without transient errors ($a_\text{trans}$), the accuracy without octave errors $a_\text{oct}$ and the adjusted accuracy without both of these errors ($a_\text{adj}$). The accuracy is calculated by dividing the number of correctly determined frames by the total number of frames with notes. The only frames without notes are positioned at the start and end of a file. We omitted these frames from the results, as they are always correctly determined to be empty frames. Erroneous notes due to transients usually have a lower amplitude where octave errors usually have a higher amplitude compared the amplitude of the base note. This allows is to distinguish between them. This information cannot be used to detect transients in real-time, as we compare the amplitude of the erroneous note to amplitude of the correct note in the following frames.
\begin{table}[H]\vspace{-3mm}
    \centering
    \begin{tabular}{r|rrrr}
        $n_\text{start}$ & $a$ & $a_\text{trans}$ & $a_\text{oct}$ & $a_\text{adj}$ \\
        \hline
        $\text{E}_2$ & 91.4\% & 95.2\% & 96.2\% & 100\% \\
        $\text{F}_2$ & 92.3\% & 96.2\% & 96.2\% & 100\% \\
        $\text{G}_2$ & 92.0\% & 95.0\% & 97.0\% & 100\% \\
        $\text{A}_2$ & 93.0\% & 94.0\% & 99.0\% & 100\% \\
        $\text{B}_2$ & 92.8\% & 96.9\% & 95.9\% & 100\% \\
        $\text{C}_3$ & 92.8\% & 92.8\% & 100\%  & 100\% \\
        $\text{D}_3$ & 96.0\% & 98.0\% & 98.0\% & 100\% \\
        $\text{E}_3$ & 96.7\% & 96.7\% & 100\%  & 100\%
    \end{tabular}
    \caption{Accuracy of monophonic $f_0$ estimation on C major scale at two notes per second for different starting positions. $a_\text{trans}$ is the accuracy without transient errors and $a_\text{oct}$ is without octave errors. $a_\text{adj}$ is the accuracy without both of these errors.}
    \label{tab:monperf}
\end{table}\vspace{-3mm}
We can see that the transient errors occur about 96\% of the time. There are two outliers at $\text{C}_3$ and $\text{D}_3$, which are the cause of inconsistent quality in the recording of the scales. This shows us that the guitarist has to play carefully for the best results. Furthermore, we noticed that the first played note almost always has a transient error. This could be because we start playing from a silent signal instead of changing from a previous note. Also, the transient errors show up more often between lower frequency notes. Almost all octave errors are at note changes, so better transient filtering will also solve these cases. Other than these errors, the system seems to work perfectly.

@@

To test the polyphonic performance of our system, we let it determine the played notes in an interval. We play every interval within an octave to test if there are any problems with separating the notes in specific intervals. For instance, some interference may cause oscillating peaks in the frequency domain. Again, we try all the intervals at different root notes, as the pitch of the root note may make a big difference in performance. As we mentioned before, when two peaks are too close together, we cannot distinguish them. Often, the second note is distinguished an octave higher. We set the overtone error threshold to 10 cents. This is a relatively small value, but this allows us to see small problems in our algorithms better. Furthermore, as shown in Table~\ref{tab:overseries}, we should not use a constant value, but use a different value for every $n$-th overtone. In Tables~\ref{tab:polperf1}--\ref{tab:polperf3} is a summary of the results. Here, we list the number of semitones $n$ between the two notes in an interval, if both notes were correctly found. If not, we denote if the second note was detected an octave higher. This often happens, because we cannot distinguish two notes in small intervals at low pitches. The last column shows if there were other notes reported which were not transients.

@@

\section{Conclusions}
In this paper, we presented and implemented a Fourier transform based AMT system and tested if these methods are feasible for real-time transcription. We divided the task in three components which can all be separately interchanged to facility specialized research in one of the components. The components are: obtaining the frequency domain representation, spectral peak picking and note selection. As shown in Section~\ref{sub:algspeed}, these tasks can be performed well within our real-time constraint of 6 ms. However, when using Fourier transform based methods, we need large frame sizes to get the required accuracy to discern lower pitched notes. Given the maximum sampling rate of widespread audio interfaces, it takes too long to fill a frame. The current implementation requires frame sizes of 16384 samples, which equates to frame lengths of 85 ms. As described in Section~\ref{sec:future}, we might still be able to use Fourier transform based methods for real-time transcription. However, other transforms such as CQT or wavelet transform might be better suited for real-time transcription, as we can still spare much compute time to lower the frame-wait time.

@@

The current implementation has no onset detection. We stated before that onset detection may help to filter out transients. If we can detect the transient in the time domain, we can discard all samples containing the transient. This may however be very difficult to do in real-time. If we can detect that there was a transient in a transformed frame, we are too late to act on it due to the large frame times. We can circumvent this by transforming partially overlapping frames. This could be done in different threads to limit the extra computational pressure which could impact our latency. However, as we are about 240 times real-time, we could likely perform it single threaded. Now, if we have a bad frame, we have to wait a full frame (85 ms) for the next estimation. However, if we use $n$ partially overlapping frames, we can try another estimation in $85/n$ ms. This will increase the responsiveness of our system greatly.
